Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?

### 2020-06-25 analysis/paper/preprint-pdf.Rmd
Plant Dis format; Split pre-print and paper Rmds for PDF and Word out

---
title: "Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?"
author: |
  | Adam H. Sparks^[Corresponding author: Adam.Sparks@usq.edu.au] \textsuperscript{1\dag}, Nancy P. Castilla \textsuperscript{1}, Bjoern Ole Sander \textsuperscript{1}
  | \textsuperscript{1}International Rice Research Institute,
  | Crop and Environmental Sciences Division, Los Baños, Laguna, Philippines, 4031
  | \textsuperscript{\dag} Current Address: University of Southern Queensland,
  | Centre for Crop Health, West Street,
  | Tooowoomba, Queensland, Australia, 4350
csl: "../templates/european-journal-of-plant-pathology.csl" # Insert path for the bib-style
abstract: |
  Water and nitrogen management play vital roles in rice production. However, the mismanagement of these two crop inputs may trigger the development of sheath blight of rice, caused by _Thanatephorus cucumeris_ (strain AG1-IA), which is favored by wet conditions, high relative humidity, and high nitrogen fertilizer levels. To understand how different combinations of water and nitrogen management affect sheath blight epidemics, we conducted two separate split--plot experiments with a water saving (alternate wetting and drying) regime and a traditional puddled regime combined with differing nitrogen treatments in the dry seasons of 2015 and 2016. Disease was scored using the same methodology in both experiments using a sheath blight assessment scale for field evaluation developed at the International Rice Research Institute to assess the severity on infected sheaths and leaves while sheath blight incidence on tillers were counted per hill. We were unable to detect any difference in the incidence of tiller sheath blight due to irrigation, tiller and leaf sheath blight did clearly differ statistically by irrigation treatment, but leaf sheath blight severity did not. Our findings indicate that farmers can adopt water saving technologies without risking increased sheath blight incidence. We suggest that further cross--disciplinary research in this area is warranted. All data generated from this project and code used for analysis are available as a reproducible research compendium from <https://doi.org/>.
bibliography: references.bib
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    #reference_docx: "../templates/template.docx"
header-includes:
  - \usepackage[T1]{fontenc}
  - \usepackage[utf8]{inputenc}
  - \DeclareUnicodeCharacter{2500}{\mbox{\vrule height2.2ptdepth-1.8ptwidth.5em}}
---

@@

knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  relative_path = TRUE,
  comment = "#>",
  fig.path = "figures/",
  fig.width = 3.31,
  dev = c("postscript", "png"),
  dpi = 300
)

@@

**Keywords** Disease $\cdot$ Sheath blight $\cdot$ *Thanatephorus cucumeris*

@@

Sheath blight disease, caused by (*Thanatephorus cucumeris* (A.B. Frank) Donk.), anastomosis group 1, of rice is an economically important disease worldwide throughout tropical and temperate production areas.
Under conducive conditions the disease can cause up to 50&nbsp;\% yield losses [@Marchetti1991].
Studies by Savary *et al.* [-@Savary2000] estimated normal yield losses in tropical irrigated lowland rice in Asia at 5 -- 10&nbsp;\% annually.
The disease is best managed through an integrated disease management approach, which includes resistance [@Kumar2009; @Srinivasachary2011], fungicide applications where affordable [@Rush1983; @Groth2008], nitrogen (N) management [@Slaton2003; @Castilla1996; @Tang2007] and water management [@Castilla1996].

Alternate wetting and drying (AWD) is an irrigation technique for rice (*Oryza sativa* L.) developed by the International Rice Research Institute (IRRI) and its partners that saves about 15 -- 40&nbsp;\% of irrigation water while not affecting yield [@Bouman2001; @Feng2007].
In AWD rice, fields are exposed to several dry phases during the growing season without exposing the plants to water stress.
In order to avoid yield decline under AWD, "safe" thresholds have been developed.
Under safe AWD irrigation water is applied when the field water level reaches 15&nbsp;cm below the soil surface [@Richards2014].
Fields are furthermore kept puddled with water standing during the flowering period to avoid plant stress.
As an added benefit to saving water, AWD also reduces greenhouse gas (GHG) emissions of rice fields, which are a substantial factor in the GHG budget of rice producing countries, by around 50 &nbsp;\% [@Yan2005; @Sander2015].

@@

### Experimental Design {-}

@@

Irrigation in AWD plots was determined by the water level in plots, *i.e.* when the water level reached 15&nbsp;cm below the soil surface irrigation water was applied to a level of 5&nbsp;cm.
In puddled plots a standing water layer of 3 -- 5&nbsp;cm was maintained throughout the growing season.

@@

### Crop Establishment {-}

@@

### Inoculation Procedure {-}

@@

### Experimental Design {-}

@@

In both experiments, two sample areas, sized 1&nbsp;m&nbsp;x&nbsp;1&nbsp;m, per plot were assessed.
The total number of tillers per hill and number of tillers with sheath blight (incidence) were measured for nine hills per sample area. Tiller sheath severity was measured for four tillers per hill and six leaves tiller using a disease assessment rating scale for field evaluation developed at IRRI.
The scale is an unequally spaced categorical scale where 0 -- no disease; 1 -- trace to 1&nbsp;\% severity; 2 -- 1 to 5&nbsp;\% severity; 3 -- 5 to 15&nbsp;\% severity; 4 -- 15 to 50&nbsp;\% severity; 5 -- 50 to 100&nbsp;\% severity.
Severity was considered to be the amount of leaf or sheath tissue covered by sheath blight lesions.
A leaf was considered living if at least 50&nbsp;% of its area was not brown or dark brown due to natural senescence.
A leaf that was yellow green or light green due to a disease, such as tungro or nitrogen deficiency was still considered a living leaf.
The same rating methodology was used in both experiments.
Five disease assessments were made in the 2015 experiment and four disease assessments were made in the 2016 experiment, respectively.

@@

### Tiller Sheath Blight Incidence {-}

@@

### Tiller Sheath Blight Severity {-}

@@

### Leaf Sheath Blight Severity {-}

@@

## The Interaction of Irrigation and N Management {-}

@@

Climate change is predicted to decrease sheath blight risk over the longer term [@Kim2015].
However, over the next 10 -- 20 years the risk is predicted to remain the same as from 2000 -- 2010 [@Kim2015].
Therefore, it is likely to remain an important disease in the near future with little resistance available in current varieties.
However, based on these studies the use of AWD technologies would not appear to be a factor that increases the incidence or severity of rice sheath blight disease.

# Notes {-}

## Reproducibility Statement {-}

@@

## Author Contributions {-}

## Acknowledgments {-}

The authors wish to thank the IRRI Biometrics group for their input in the design of the experiments and Dr. R. King and Mr. D. Burrell of the University of Southern Queensland Statistical Consulting Unit for their insightful comments on the analysis of this paper.
We would also like to thank Mr. D. Amozola, Mr. P. Escandor and Mr. P. Lapis for their technical support in the field.
Lastly, we would like to remember our team mate Michael Noel, an integral part of our plant disease management team at IRRI and who helped design and managed the first season of the project but passed away in September 2016.

# References {-}
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document: -->

>>>>

---
title: "Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?"
author: |
  | Adam H. Sparks^[Corresponding author: Adam.Sparks@usq.edu.au] \textsuperscript{1\dag}, Nancy P. Castilla \textsuperscript{1}, Bjoern Ole Sander \textsuperscript{1}
  | \textsuperscript{1}International Rice Research Institute,
  | Crop and Environmental Sciences Division, Los Baños, Laguna, Philippines, 4031
  | \textsuperscript{\dag} Current Address: University of Southern Queensland,
  | Centre for Crop Health, West Street,
  | Tooowoomba, Queensland, Australia, 4350
csl: "../templates/plant_disease.csl" # Insert path for the bib-style
abstract: |
  Water and nitrogen management play vital roles in rice production. However, the mismanagement of these two crop inputs may trigger the development of sheath blight of rice, caused by _Thanatephorus cucumeris_ (strain AG1-IA), which is favored by wet conditions, high relative humidity, and high nitrogen fertilizer levels. To understand how different combinations of water and nitrogen management affect sheath blight epidemics, we conducted two separate split--plot experiments with a water saving (alternate wetting and drying) regime and a traditional puddled regime combined with differing nitrogen treatments in the dry seasons of 2015 and 2016. Disease was scored using the same methodology in both experiments using a sheath blight assessment scale for field evaluation developed at the International Rice Research Institute to assess the severity on infected sheaths and leaves while sheath blight incidence on tillers were counted per hill. We were unable to detect any difference in the incidence of tiller sheath blight due to irrigation, tiller and leaf sheath blight did clearly differ statistically by irrigation treatment, but leaf sheath blight severity did not. Our findings indicate that farmers can adopt water saving technologies without risking increased sheath blight incidence. We suggest that further cross--disciplinary research in this area is warranted.
bibliography: references.bib
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    #reference_docx: "../templates/template.docx"
header-includes:
  - \usepackage[T1]{fontenc}
  - \usepackage[utf8]{inputenc}
  - \DeclareUnicodeCharacter{2500}{\mbox{\vrule height2.2ptdepth-1.8ptwidth.5em}}
---

@@

knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  relative_path = TRUE,
  comment = "#>",
  fig.path = "figures/",
  fig.width = 3.25,
  dev = c("postscript", "png"),
  dpi = 300
)

@@

**Keywords** Rice $\cdot$ Water Management $\cdot$ Nitrogen Management $\cdot$ Sheath Blight $\cdot$ *Thanatephorus cucumeris*

@@

Sheath blight disease, caused by (*Thanatephorus cucumeris* (A.B. Frank) Donk.), anastomosis group 1, of rice is an economically important disease worldwide throughout tropical and temperate production areas.
Under conducive conditions the disease can cause up to 50&nbsp;\% yield losses [@Marchetti1991].
Studies by Savary *et al.* [-@Savary2000] estimated normal yield losses in tropical irrigated lowland rice in Asia at 5 to 10&nbsp;\% annually.
The disease is best managed through an integrated disease management approach, which includes resistance [@Kumar2009; @Srinivasachary2011], fungicide applications where affordable [@Rush1983; @Groth2008], nitrogen (N) management [@Slaton2003; @Castilla1996; @Tang2007] and water management [@Castilla1996].

Alternate wetting and drying (AWD) is an irrigation technique for rice (*Oryza sativa* L.) developed by the International Rice Research Institute (IRRI) and its partners that saves about 15 to 40&nbsp;\% of irrigation water while not affecting yield [@Bouman2001; @Feng2007].
In AWD rice, fields are exposed to several dry phases during the growing season without exposing the plants to water stress.
In order to avoid yield decline under AWD, "safe" thresholds have been developed.
Under safe AWD irrigation water is applied when the field water level reaches 15&nbsp;cm below the soil surface [@Richards2014].
Fields are furthermore kept puddled with water standing during the flowering period to avoid plant stress.
As an added benefit to saving water, AWD also reduces greenhouse gas (GHG) emissions of rice fields, which are a substantial factor in the GHG budget of rice producing countries, by around 50 &nbsp;\% [@Yan2005; @Sander2015].

@@

## 2015 Experimental Design {-}

@@

Irrigation in AWD plots was determined by the water level in plots, *i.e.* when the water level reached 15&nbsp;cm below the soil surface irrigation water was applied to a level of 5&nbsp;cm.
In puddled plots a standing water layer of 3 to 5&nbsp;cm was maintained throughout the growing season.

@@

@@

@@

## 2016 Experimental Design {-}

@@

In both experiments, two sample areas, sized 1&nbsp;m&nbsp;x&nbsp;1&nbsp;m, per plot were assessed.
The total number of tillers per hill and number of tillers with sheath blight (incidence) were measured for nine hills per sample area. Tiller sheath severity was measured for four tillers per hill and six leaves tiller using a disease assessment rating scale for field evaluation developed at IRRI.
The scale is an unequally spaced categorical scale where (i) 0 -- no disease; (ii) 1 -- trace to 1&nbsp;\% severity; (iii) 2 -- 1 to 5&nbsp;\% severity; (iv) 3 -- 5 to 15&nbsp;\% severity; (v) 4 -- 15 to 50&nbsp;\% severity; and (vi) 5 -- 50 to 100&nbsp;\% severity.
Severity was considered to be the amount of leaf or sheath tissue covered by sheath blight lesions.
A leaf was considered living if at least 50&nbsp;% of its area was not brown or dark brown due to natural senescence.
A leaf that was yellow green or light green due to a disease, such as tungro or nitrogen deficiency was still considered a living leaf.
The same rating methodology was used in both experiments.
Five disease assessments were made in the 2015 experiment and four disease assessments were made in the 2016 experiment, respectively.

@@

@@

@@

@@

@@

Climate change is predicted to decrease sheath blight risk over the longer term [@Kim2015].
However, over the next 10 to 20 years the risk is predicted to remain the same as from 2000 to 2010 [@Kim2015].
Therefore, it is likely to remain an important disease in the near future with little resistance available in current varieties.
However, based on these studies the use of AWD technologies would not appear to be a factor that increases the incidence or severity of rice sheath blight disease.

# Acknowledgments {-}

The authors wish to thank the IRRI Biometrics group for their input in the design of the experiments and Dr. R. King and Mr. D. Burrell of the University of Southern Queensland Statistical Consulting Unit for their insightful comments on the analysis of this paper.

We would also like to thank Dr. J. Villa, Mr. B. Macasero, Mr. D. Amozola, Mr. P. Escandor and Mr. P. Lapis for their technical support in the field.

Funding for this research was provided through the CGIAR Research Program on rice agri-food systems Global Rice Science Partnership (GRiSP).

Lastly, we would like to remember our team mate Michael Noel, an integral part of our plant disease management team at IRRI and who helped design and managed the first season of the project but passed away in September 2016.

# Literature Cited {-}
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document: -->

# Author-Recommended Internet Resources {-}

@@
